Data Descriptor: Corresponding author journal publishing activities 2014-2018

### 2019-10-21 analysis/paper.Rmd
addressing @Ahobert comments and questions #1

There have also been analytical applications for such datasets. Major bibliometric databases including the Web of Science, Scopus and Dimension have started to integrate open access status information. In these databases, Unpaywall data is often used to identify openly available full-texts at the article-level, while the Directory of Open Access Journals provides information about if a journal is fully open access. In its most recent edition, the Leiden Ranking presents open access indicators at the level of universities [@leiden]. In addition to university-specific data, the German Open Access Monitor provides breakdowns by publisher and journal and includes publications from non-university research institutions based in Germany [@oa_monitor].

@@

Following @Palzenberger_2015, data were retrieved from the in-house Web of Science database maintained by the German Competence Center for Bibliometrics. The following data were obtained and aggregated:

- Web of Science collections SCI, SSCI and AHCI <!-- why only these collections?-->
- Article types Original Articles and Reviews
- Countries of affiliation
- ISSN
- Publication years 2014 - 2018
- Author roles, interpreting the role "reprint author" as corresponding authorship.

@@

Indeed, the Web of Science has begun to systematically keep track of more than one corresponding author recently: <!-- maybe stress that before 2016 only one unique corresponding author was allowed?--> We found that the Web of Science did record more than one corresponding author for 6.2% of indexed articles published between 2014 - 2018 (see Figure 1). Moreover, a 5% of corresponding authors author was internationally co-located (see Figure 2). We therefore<!--delete therefore?--> decided to use whole counting where every original article or review was counted once per country of affiliation of the corresponding author(s).

@@

After obtaining aggregated article counts at the level of countries and journals from the Web of Science, Crossref’s REST API was queried using the rcrossref package [@rcrossref] to retrieve publisher <!--'publisher names' or 'publishers'--> and journal titles. To control for developments of the publishing market resulting in name changes of publishers or journal titles over time, only the most frequent field name was used from the API facet counts. <!--most frequent name amongst all considered articles or in the whole of crossref? for this specific journal or always the same name over all journals?--> Drawing on ISSN-L, a journal identifier that links between different media versions of a journal, ISSN variants were obtained to improve the retrieval. Next, the open access status per journal was identified and added to the dataset using the ISSN-Gold-OA 3.0 list [@issn]. Again, ISSN-L were used for the matching procedure.

@@

The first file, named `journal_publisher_14_18`, contains information about the publishing market landscape for the period 2014-2018 in terms of the aggregated number of articles and reviews  by year and journal represented by the ISSN identifier indexed in the Web of Science. Table 1 describes the data variables in detail. We were able to retrieve publisher and journal title from Crossref for `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow() / publisher_df %>% distinct(issn_l) %>% nrow() * 100` %
 of all journals, representing `r publisher_df %>% filter(!is.na(publisher)) %>% .$articles %>% sum() / sum(publisher_df$articles) * 100` % of the overall article volume. `r publisher_df %>% filter(oa_journal == TRUE) %>% distinct(issn_l) %>% nrow()` out of `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow()`
 investigated journals with `ISSN-L` were identified as fully open access journals. In total, the dataset provides information for `r sum(publisher_df$articles)` records indexed in the Web of Science with journal information in the five-years period 2014-2018.
 <!-- are only journals with valid issn considered? Are there journals in the dataset without issn-l?-->
<!--what about the 9% that could not be matched with crossref? are they excluded or included with WoS information /without journal information?-->
Data Schema `journal_publisher_14_18`: <!--mention the abbreviation 'KB Web of Science' before somewhere when talking about KB-->

@@

Although the Web of Science database is a well-established bibliometric database it has certain limitations that need careful consideration when using and interpreting our data. First, the Web of Science is selective in terms of journal inclusion and is not covering the whole journal portfolio of the investigated publishers. Although publisher names were disambiguated, journal title transfers between publishers were not tracked. Furthermore, the year of inclusion in a journal issue, i.e. print publication date, was used. However, there can be a considerable time-lag between online-first publication and issue inclusion. <!--mention also selection bias of WoS in terms of countries, ie. publications from EU, northern america much better covered than other continents-->

@@

The ISSN GOLD-OA list <!--add version ?--> [@issn] was used to determine fully open access journals. However, it does not contain information when a subscription-based journal became fully open access. The figures should therefore be interpreted as numbers of articles available in current fully open access journals, instead of published in a fully open access journal at the time of publication. Open access article counts are therefore upper estimates. <!-- Is it really upper estimates? What about reverse flip?-->

In the end, we present a novel bibliometric dataset representing the journal publishing market with a particular focus on open access. Although access to the underlying data infrastructure is restricted, we decided to openly share the aggregated dataset as part of a research compendium to demonstrate <!-- ensure instead of demonstrate?--> the technical reproducibility of this work.

>>>>

There have also been analytical applications for such datasets. Major bibliometric databases including the Web of Science, Scopus and Dimension have started to integrate open access status information. In these databases, Unpaywall data is often used to identify openly available full-texts at the article-level, while the Directory of Open Access Journals provides information about if a journal is fully open access. In its most recent edition, the Leiden Ranking presents open access indicators at the level of universities [@leiden]. In addition to university-specific data, the German Open Access Monitor provides breakdowns by publisher and journal and includes publications from non-university research institutions based in Germany [@oa_monitor]. In its [Publisher Analytics Reports](https://clarivate.com/webofsciencegroup/solutions/publisher-analytics-reports/) Clarivate Analtics presents the performance of publishers across subjects, institutions and countries based on Web of Science data.

@@

Following @Palzenberger_2015, data were retrieved from the in-house Web of Science database maintained by the German Competence Center for Bibliometrics (KB Web of Science). The following data were obtained and aggregated:

- Web of Science core collections SCI, SSCI and AHCI 
- Article types Original Articles and Reviews
- Countries of affiliation
- ISSN
- Publication years 2014 - 2018
- Author roles, interpreting the role "reprint author" as corresponding authorship.

@@

Indeed, the Web of Science has begun to systematically keep track of more than one corresponding author recently: We found that the Web of Science did record more than one corresponding author for 6.2% of indexed articles published between 2014 - 2018. However, Figure 1 suggests that indexing of more than one corresponding author started in 2016. Moreover, a 5% of corresponding authors author was internationally co-located (see Figure 2). We therefore decided to use whole counting where every original article or review was counted once per country of affiliation of the corresponding author(s).

@@

After obtaining aggregated article counts at the level of countries and journals from the Web of Science, Crossref’s REST API was queried using the rcrossref package [@rcrossref] to retrieve publishers and journal titles. To control for developments of the publishing market resulting in name changes of publishers or journal titles over time, only the most frequent field name for the specific journal was used from the API facet counts. Drawing on ISSN-L, a journal identifier that links between different media versions of a journal, ISSN variants were obtained to improve the retrieval. Next, the open access status per journal was identified and added to the dataset using the ISSN-Gold-OA 3.0 list [@issn]. Again, ISSN-L were used for the matching procedure.

@@

The first file, named `journal_publisher_14_18`, contains information about the publishing market landscape for the period 2014-2018 in terms of the aggregated number of articles and reviews  by year and journal represented by the ISSN identifier indexed in the Web of Science. Table 1 describes the data variables in detail. We were able to retrieve publisher and journal title from Crossref for `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow() / publisher_df %>% distinct(issn_l) %>% nrow() * 100` %
 of all journals, representing `r publisher_df %>% filter(!is.na(publisher)) %>% .$articles %>% sum() / sum(publisher_df$articles) * 100` % of the overall article volume. `r publisher_df %>% filter(oa_journal == TRUE) %>% distinct(issn_l) %>% nrow()` out of `r publisher_df %>% filter(!is.na(publisher)) %>% distinct(issn_l) %>% nrow()`
 investigated journals with `ISSN-L` were identified as fully open access journals. In total, the dataset provides information for `r sum(publisher_df$articles)` records indexed in the Web of Science with journal information in the five-years period 2014-2018..
 <!-- are only journals with valid issn considered? Are there journals in the dataset without issn-l?-->
<!--what about the 9% that could not be matched with crossref? are they excluded or included with WoS information /without journal information?-- per Wos ISSN-->
Data Schema `journal_publisher_14_18`:

@@

Although the Web of Science database is a well-established bibliometric database it has certain limitations that need careful consideration when using and interpreting our data. First, the Web of Science is selective in terms of journal inclusion and is not covering the whole journal portfolio of the investigated publishers. There is also evidence about a considerable indexing lag of non-English journal publications relevant to domestic research [@Lopez_Pineiro_2014]. Although publisher names were disambiguated, journal title transfers between publishers were not tracked. Furthermore, the year of inclusion in a journal issue, i.e. print publication date, was used. However, there can be a considerable time-lag between online-first publication and issue inclusion.

@@

The ISSN GOLD-OA list [@issn] was used to determine fully open access journals. However, it does not contain information when a subscription-based journal became fully open access. The figures should therefore be interpreted as numbers of articles available in current fully open access journals, instead of published in a fully open access journal at the time of publication. Open access article counts are therefore upper estimates. <!-- Is it really upper estimates? What about reverse flip?-->

In the end, we present a novel bibliometric dataset representing the journal publishing market with a particular focus on open access. Although access to the underlying data infrastructure is restricted, we decided to openly share the aggregated dataset as part of a research compendium to ensure the technical reproducibility of this work.
